---
title: "Worked example: simulated beetle clade"
author: "Heath Blackmon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked example: simulated beetle clade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette walks through a complete chromePlus analysis end-to-end on
a non-trivial dataset: a simulated 100-species beetle clade where
chromosome number evolves jointly with a binary sex-chromosome-system
trait (standard vs. fused). The whole analysis runs in under a minute on
a modern laptop.

The data are simulated (see `?beetleChrom`) and ship with the package so
you don't need to download anything.

## Load the data

```{r load}
library(chromePlus)
library(diversitree)
library(ape)

data(beetleChrom)
tree <- beetleChrom$tree
dat  <- beetleChrom$data

# Ntip and counts
ape::Ntip(tree)
table(dat$state)
table(dat$chrom)
```

A quick look:

```{r peek}
head(dat)
```

## Build the probability matrix

`datatoMatrix()` turns the tidy data frame into the probability matrix
that `diversitree` expects. Because we want a binary trait in the
model we use `hyper = TRUE`, and we pass `state.names` so every
downstream object (columns of the matrix, parameter names of the fit)
is self-documenting — no more guessing whether "asc1" is the rate in
the standard or fused state.

```{r datamat}
dat.mat <- datatoMatrix(
  x     = dat[, c("species", "chrom", "prob")],
  hyper = TRUE,
  state.names = c("standard", "fused")
)
dim(dat.mat)
colnames(dat.mat)[c(1:4, (ncol(dat.mat) - 3):ncol(dat.mat))]
```

The left half of the columns is the "standard" state; the right half
carries the `.fused` suffix because we asked for descriptive names.

## Fit an mkn chromosome-evolution model

```{r fit-mkn, cache = TRUE}
lik <- make.mkn(tree, states = dat.mat, k = ncol(dat.mat),
                strict = FALSE, control = list(method = "ode"))

con.lik <- constrainMkn(
  data  = dat.mat,
  lik   = lik,
  hyper = TRUE,
  state.names = c("standard", "fused"),
  constrain = list(drop.demi = TRUE, drop.poly = TRUE)
)

argnames(con.lik)
```

Thanks to `state.names`, the returned parameters are `asc.standard`,
`asc.fused`, `desc.standard`, `desc.fused`, `tran.standard.to.fused`,
and `tran.fused.to.standard`. Nothing is named `asc1` or `asc2` and
nothing is numerically ambiguous.

```{r mle, cache = TRUE}
set.seed(1)
sv <- setNames(startVals(n = length(argnames(con.lik)),
                         min = 0.01, max = 0.1),
               argnames(con.lik))
fit <- find.mle(con.lik, sv, method = "subplex",
                control = list(maxit = 2000))
round(coef(fit), 4)
```

The rate names in `coef(fit)` are the same descriptive labels, so you
can read the fitted values directly: how fast chromosome number goes
up in each state, how fast it goes down, and how fast the
sex-chromosome-system transitions happen in each direction.

## Visualise the tree coloured by state

```{r plot, fig.height = 7}
state_col <- ifelse(dat$state[match(tree$tip.label, dat$species)] ==
                    "standard", "steelblue", "tomato")
plot(tree, tip.color = state_col, cex = 0.4, no.margin = TRUE)
legend("bottomleft",
       legend = c("standard", "fused"),
       text.col = c("steelblue", "tomato"),
       bty = "n")
```

## What to try next

* Fit the same data under `constrainMuSSE()` to let speciation and
  extinction depend on the binary state.
* Re-fit with `constrain = list(nometa = TRUE)` to ask whether the
  state actually matters for chromosome evolution — compare likelihoods.
* Swap your real data into `datatoMatrix()` with the same `state.names`
  convention.
