---
title: "Fast Mkn Likelihoods for Large Chromosome State Spaces"
author: "Heath Blackmon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fast Mkn Likelihoods for Large Chromosome State Spaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## When to use this

The default chromePlus pipeline (`make.mkn()` from diversitree, then
`constrainMkn()`) integrates a dense ODE for every branch of the tree. At small
state-space sizes that is fast, but cost grows roughly as $k^2$ where $k$ is
the number of states (chromosome counts × hyperstates). Once $k \geq 100$,
each likelihood evaluation can take tens to hundreds of milliseconds, which
becomes painful inside an MCMC of 50,000+ steps or across many model variants.

The chromosome rate matrix $Q$ is extremely sparse — only ~6 non-zeros per row
regardless of $k$ (ascending, descending, polyploidy, two demiploidy entries,
hyperstate transition). `make.mkn.fast()` exploits this with a sparse
**uniformization-based** Felsenstein pruning kernel implemented in C++. The
result is the same likelihood, computed 30-240× faster at $k = 100$ to 600.

This vignette shows two things:

1. **Identical output.** `make.mkn.fast()` produces log-likelihoods that match
   diversitree's `make.mkn() |> constrainMkn()` to ~1e-8 across a range of
   problem sizes and constraint settings.
2. **Speed.** The fast path is dramatically faster, especially at large $k$.

## Drop-in usage

`make.mkn.fast()` takes the same `tree`, data matrix, `hyper`, `polyploidy`,
`oneway`, `state.names`, and `constrain` arguments as the
`make.mkn() |> constrainMkn()` chain, and returns a function with the same
calling convention:

```{r usage, eval = FALSE}
library(chromePlus); library(diversitree); library(ape)

# Suppose dat.mat came from datatoMatrix(...).
ll_fast <- make.mkn.fast(tree, dat.mat, hyper = TRUE,
                         constrain = list(drop.demi = TRUE))

# Same parameter names and order as constrainMkn()
argnames(ll_fast)

# Evaluate at a parameter vector
pars <- setNames(rep(0.05, length(argnames(ll_fast))), argnames(ll_fast))
ll_fast(pars)

# Works directly with find.mle and mcmc
fit   <- find.mle(ll_fast, pars, method = "subplex")
chain <- diversitree::mcmc(ll_fast, pars, nsteps = 10000, w = 0.05,
                            prior = function(p) sum(dexp(p, rate = 10, log = TRUE)),
                            lower = 1e-6, print.every = 0)
```

## Build a reproducible test problem

We simulate phylogenetically-correlated chromosome data so that sibling
species have similar karyotypes, as in real data. This is necessary for the
likelihood to be informative (random-uniform chromosome assignments produce
disjoint partial-likelihood supports at sibling joins, which is unrealistic).

```{r build}
suppressPackageStartupMessages({
  library(chromePlus); library(diversitree); library(ape)
})

build_problem <- function(k_target, n_tips = 100, seed = 1) {
  half <- k_target / 2
  chrom_range <- c(5, 5 + half - 1)
  set.seed(seed)
  tree <- rcoal(n_tips, tip.label = paste0("sp", 1:n_tips))
  sim <- suppressMessages(simChrom(
    tree = tree, model = "ChromPlus",
    pars = c(asc1 = 0.10, asc2 = 0.12, desc1 = 0.08, desc2 = 0.09,
             dem1 = 0,    dem2 = 0,    pol1  = 0.005, pol2  = 0.005,
             tran12 = 0.02, tran21 = 0.02,
             root.chrom = round(mean(chrom_range)), root.state = 0),
    limits = chrom_range))
  chroms <- pmin(pmax(as.numeric(sim$chrom.num), chrom_range[1]),
                 chrom_range[2])
  bin <- as.numeric(sim$binary.state)
  dat <- data.frame(species = tree$tip.label, chrom = chroms,
                    prob = ifelse(bin == 0, 1, 0))
  dat.mat <- datatoMatrix(x = dat, range = chrom_range, hyper = TRUE)
  list(tree = tree, dat.mat = dat.mat, k = k_target)
}
```

## Identical output

We construct both pipelines and compare log-likelihoods at several parameter
vectors. Differences should be at the level of ODE-solver accuracy
(`~1e-8`), not larger.

```{r identical}
prob <- build_problem(k_target = 80, n_tips = 50)

# diversitree path
lik <- suppressMessages(make.mkn(prob$tree, states = prob$dat.mat,
                                  k = ncol(prob$dat.mat), strict = FALSE,
                                  control = list(method = "ode", tol = 1e-10)))
con <- suppressMessages(constrainMkn(data = prob$dat.mat, lik = lik,
                                      hyper = TRUE,
                                      constrain = list(drop.demi = TRUE)))

# chromePlus fast path
ll_fast <- make.mkn.fast(prob$tree, prob$dat.mat, hyper = TRUE,
                         constrain = list(drop.demi = TRUE))

# Same parameters, same order
argn <- argnames(ll_fast)
stopifnot(setequal(argn, diversitree::argnames(con)))

# Evaluate at five random parameter draws and compare
set.seed(42)
diffs <- replicate(5, {
  pars <- setNames(runif(length(argn), 0.001, 0.2), argn)
  abs(con(pars[diversitree::argnames(con)]) - ll_fast(pars))
})
diffs
max(diffs)
```

The differences are well below 1e-7. Note: the residual is on diversitree's
side — the ODE accumulates numerical error proportional to its tolerance,
while uniformization is exact to machine precision. Tighten the ODE tolerance
(`control = list(tol = 1e-12)`) and the agreement gets even closer.

## Speed comparison

```{r speed}
bench_at <- function(k_target, n_tips = 100, n_reps = 20, seed = 1) {
  prob <- build_problem(k_target, n_tips, seed)
  lik <- suppressMessages(make.mkn(prob$tree, states = prob$dat.mat,
                                    k = ncol(prob$dat.mat), strict = FALSE,
                                    control = list(method = "ode", tol = 1e-8)))
  con <- suppressMessages(constrainMkn(data = prob$dat.mat, lik = lik,
                                        hyper = TRUE,
                                        constrain = list(drop.demi = TRUE)))
  ll_fast <- make.mkn.fast(prob$tree, prob$dat.mat, hyper = TRUE,
                            constrain = list(drop.demi = TRUE))
  argn <- argnames(ll_fast)
  pars <- setNames(rep(0.05, length(argn)), argn)
  if ("pol1" %in% argn)   pars["pol1"]   <- 0.001
  if ("pol2" %in% argn)   pars["pol2"]   <- 0.0015
  if ("tran12" %in% argn) pars["tran12"] <- 0.005
  if ("tran21" %in% argn) pars["tran21"] <- 0.004

  v_div  <- con(pars[diversitree::argnames(con)])
  v_fast <- ll_fast(pars)
  t_div  <- system.time(for (i in seq_len(n_reps)) v_div  <- con(pars[diversitree::argnames(con)]))[["elapsed"]] / n_reps
  t_fast <- system.time(for (i in seq_len(50 * n_reps)) v_fast <- ll_fast(pars))[["elapsed"]] / (50 * n_reps)

  data.frame(k = k_target,
             n_tips = n_tips,
             diversitree_s = signif(t_div, 3),
             fast_s        = signif(t_fast, 3),
             speedup       = signif(t_div / t_fast, 3),
             ll_diff       = signif(abs(v_div - v_fast), 3))
}

# Sweep at moderate sizes (kept small here so the vignette renders quickly)
do.call(rbind, lapply(c(50, 100, 200), bench_at, n_tips = 100))
```

On a typical workstation we observe (`n_tips = 150`, optimised installed
package):

| `k`  | diversitree | `make.mkn.fast` | speedup |
|------|-------------|------------------|---------|
| 50   | 5.8 ms      | 0.17 ms          | 34×     |
| 100  | 13 ms       | 0.32 ms          | 40×     |
| 200  | 39 ms       | 0.64 ms          | 61×     |
| 400  | 168 ms      | 1.26 ms          | 134×    |
| 600  | 472 ms      | 1.96 ms          | 241×    |

For a 100,000-step MCMC at `k = 600`: ~13 hours via diversitree drops to
~3 minutes via `make.mkn.fast()`.

## Use with `find.mle` and `mcmc`

The fast likelihood is a drop-in replacement for any diversitree-style
`function(pars)` likelihood. Both `find.mle()` and `mcmc()` work without any
changes:

```{r mle, results = "hide"}
prob <- build_problem(k_target = 50, n_tips = 50)
ll_fast <- make.mkn.fast(prob$tree, prob$dat.mat, hyper = TRUE,
                         constrain = list(drop.demi = TRUE,
                                          drop.poly = TRUE))
init <- setNames(rep(0.05, length(argnames(ll_fast))), argnames(ll_fast))
fit <- find.mle(ll_fast, init, method = "subplex",
                control = list(reltol = 1e-4, maxit = 5000))
```

```{r mle-show}
fit$lnLik
fit$par
```

```{r mcmc, results = "hide"}
chain <- diversitree::mcmc(
  ll_fast, init, nsteps = 200, w = 0.05,
  prior = function(p) sum(dexp(p, rate = 10, log = TRUE)),
  lower = 1e-6, print.every = 0)
```

```{r mcmc-show}
head(chain)
```

## Caveats

- **Mkn only.** `make.mkn.fast()` covers the `make.mkn() |> constrainMkn()`
  path. The MuSSE case (`constrainMuSSE()`, with state-dependent speciation
  and extinction) is not yet implemented because the coupled E/D equations
  do not fit a sparse-uniformization kernel cleanly. For MuSSE, continue
  using `constrainMuSSE()` for now.
- **Tip data convention.** Probabilistic tip states (rows of `dat.mat` that
  are not 0/1) are accepted; the convention follows `datatoMatrix()`'s
  output. For `make.mkn`, diversitree itself requires deterministic tips —
  if you have probabilistic tips, you were already routed to MuSSE.
- **Numerical agreement.** Differences from `constrainMkn` are on the order
  of `1e-8` and reflect diversitree's ODE tolerance, not bugs in the fast
  path. Tightening `control$tol` reduces the gap.

## How it works

`make.mkn.fast()` builds the same parameter-mapping matrix (`parMat`) as
`constrainMkn()` — the two functions share an internal helper, so model
construction is bit-identical. The difference is the per-evaluation kernel:

- **diversitree** integrates the matrix ODE
  $\frac{dY}{dt} = QY$ once per unique branch length (matrix-valued, $k^2$
  state variables), via a dense GSL `gemm` in C, then does dense pruning.
- **`make.mkn.fast()`** does direct sparse Felsenstein pruning. For each
  branch it computes $\exp(Q t) \cdot D$ via Jensen's uniformization
  (Poisson-weighted powers of $P = I + Q/\alpha$), exploiting the ~6
  non-zeros per row of $Q$. Each call is $\mathcal{O}(\mathrm{nnz}(Q) \cdot
  \alpha t)$ rather than $\mathcal{O}(k^3)$.

The savings grow with $k$ because the dense path scales superlinearly with
state-space size while the sparse path scales with the (fixed) per-row
non-zero count.
